---
title: "Segregation Indices (calculated from block groups); African American pop"
author: "Thea Rossman"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(tidycensus)

# Parameters

file_race_ethnicity_block_grps <- 
  "../data/appalachian_block_groups_race.csv"
file_centr_race_ethnicity_block_grps <-
  "../data/central_appalachian_block_groups_race.csv"
file_race_ethnicity_counties <- 
  "../data/appalachian_counties_race.csv"

acs5year <- 2018

#===============================================================================

# Code
```


```{r}
race_ethnicity_block_grps <- 
  file_race_ethnicity_block_grps %>%
  read_csv(
    col_types = 
      cols(
        census_block_fips = col_character(),
        name = col_character(),
        prop_nonwhite = col_double(),
        .default = col_integer()
      )
  )

race_ethnicity_counties <- 
  file_race_ethnicity_counties %>%
  read_csv(
    col_types = 
      cols(
        county_name = col_character(),
        state_name = col_character(),
        .default = col_integer()
      )
  )
```

Notes on this document: 

Most of the code below is copied and pasted from the "tracts" EDA. If this were production-quality code, I might make a function and think about how to do this more smoothly and with fewer lines, but this was more time efficient. 

I could not get the numbers to work for Hispanic/Latino populations and Native American populations. 


# isolation index

```{r}
# for an index to be counted, 
# counties must contain at least one census tract with prop of population of 
# interest > prop, or must have a total prop of population > prop 
  prop <- 0.05

# filter to counties with at least one tract > prop AA population
counties <- 
  c(
    race_ethnicity_block_grps %>%
      filter(black / total_population > prop) %>%
      pull(county_fips),
    race_ethnicity_counties %>%
      filter(black / total_population > prop) %>%
      pull(fips)
  ) %>%
  unique()
  
race_counties_aa_isolation_index <- 
  race_ethnicity_block_grps %>%
  select(county_fips, name, persons_i = total_population, aa_i = black) %>%
  filter(county_fips %in% counties, persons_i > 0) %>%
  group_by(county_fips) %>%
  summarize(
    aa_total = sum(aa_i),
    pop_total = sum(persons_i),
    prop_aa_county = aa_total / pop_total,
    numerator_1 = sum((aa_i / aa_total) * (aa_i / persons_i))
  ) %>%
  mutate(
    aa_isolation_index = (numerator_1 - prop_aa_county) / (1 - prop_aa_county)
  ) %>% 
  select(county_fips, aa_isolation_index)

race_counties_isolation_indices_block_grps <- 
  race_ethnicity_counties %>% 
  rename(county_fips = fips) %>% 
  left_join(race_counties_aa_isolation_index, by = "county_fips")

race_counties_isolation_indices_block_grps
```

# Dissimilarity

```{r}
counties <- 
  c(
    race_ethnicity_block_grps %>%
      filter(black / total_population > prop) %>%
      pull(county_fips),
    race_ethnicity_counties %>%
      filter(black / total_population > prop) %>%
      pull(fips)
  ) %>%
  unique()

aa_counties_dissimilarity_index <- 
  race_ethnicity_block_grps %>%
  select(
    county_fips, 
    name, 
    persons_i = total_population, 
    aa_i = black
  ) %>%
  filter(county_fips %in% counties) %>%
  group_by(county_fips) %>%
  summarize(
    aa_total = sum(aa_i),
    aa_c_total = sum(persons_i) - sum(aa_i),
    aa_dissimilarity_index = 
      0.5*sum(abs(aa_i / aa_total - (persons_i - aa_i) / aa_c_total))
  ) %>% 
  select(county_fips, aa_dissimilarity_index)

race_counties_dissimilarity_indices_block_grps <- 
  race_ethnicity_counties %>% 
  rename(county_fips = fips) %>%
  left_join(aa_counties_dissimilarity_index, by = "county_fips")
```


```{r}
race_counties_segregation_indices <- 
  race_counties_dissimilarity_indices_block_grps %>%
  left_join(
    race_counties_isolation_indices_block_grps %>% 
      select(
        county_fips, 
        aa_isolation_index
      ), 
    by = "county_fips"
  )
```

```{r}
race_counties_segregation_indices
```


```{r}
counties_segregation_geo <- 
  race_counties_segregation_indices %>%
  left_join(
    ussf::boundaries(geography = "county") %>%
      transmute(county_fips = as.integer(GEOID), geometry),
    by = "county_fips"
  )
  
plot_segregation <- function(tibble, group, index) {
  tibble %>%
    ggplot(aes(geometry = geometry, fill = index)) + 
    geom_sf() + 
    scale_fill_viridis_c() +
    theme_void() + 
    labs(
      title = str_glue(index, " of ", group, " Population"),
      subtitle = 
        str_glue("Counties with at least one census block having ", 
                 prop*100,
                 "% or more\n",
                 group, 
                 " population, or with overall\n", 
                 group, 
                 " population at least ", 
                 prop*100, 
                 "%."
                )
    )
}

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = aa_dissimilarity_index,
      geometry
    ),
  group = "African American",
  index = "Dissimilarity Index"
)

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = aa_isolation_index,
      geometry
    ),
  group = "African American",
  index = "Isolation Index"
)

```


```{r}
segregation_indices_pivoted <- 
  race_counties_segregation_indices %>%
  select(
    county_fips,
    county_name,
    state_name,
    total_population,
    black,
    isolation_black = aa_isolation_index,
    dissimilarity_black = aa_dissimilarity_index,
  ) %>%
  pivot_longer(
    cols = c(isolation_black, dissimilarity_black),
    names_to = c("index_name", "race_ethn"),
    names_sep = "_",
    values_to = "index_value"
  ) %>%
  drop_na() %>%
  mutate(
    pop_group = 
      case_when(
        race_ethn == "black" ~ black
      ),
    prop_of_population = pop_group / total_population
  )


segregation_indices_pivoted %>%
  ggplot(aes(index_value, color = index_name)) +
  geom_density() +
  facet_grid(cols = vars(race_ethn))

segregation_indices_pivoted %>%
  ggplot(aes(index_value, color = fct_reorder(race_ethn,index_value))) +
  geom_density() +
  facet_grid(cols = vars(index_name)) +
  theme(legend.position = "top") +
  labs(
    x = "Index Value (1 indicated higher segregation)",
    y = "Density (Number of Counties)",
    color = ""
  )
```


```{r}
segregation_indices_pivoted %>%
  sample_frac() %>%
  ggplot(aes(index_value, total_population, color = race_ethn)) +
  geom_point(alpha = 0.8) +
  facet_grid(cols = vars(index_name)) +
  theme(legend.position = "top") +
  labs(
    x = "Index Value (1 indicated higher segregation)",
    y = "Total Population of the County",
    color = ""
  )
```

```{r}
segregation_indices_pivoted %>%
  sample_frac() %>%
  ggplot(aes(index_value, prop_of_population, color = race_ethn)) +
  geom_point(alpha = 0.8) +
  facet_grid(cols = vars(index_name)) +
  theme(legend.position = "top") +
  guides(color = "none") +
  labs(
    title = "African American Population: Appalachia",
    x = "Index Value (1 indicated higher segregation)",
    y = "Group's Proportion of County Population",
    color = ""
  )
```


```{r, fig.asp=1}
segregation_indices_pivoted %>%
  filter(race_ethn == "black") %>%
  ggplot(aes(index_value, color = index_name)) + 
  geom_density() +
  facet_wrap(vars(state_name)) +
  theme(legend.position = "top") +
  labs(title = "Af. Am population")
```


