---
title: "poor white people live everywhere; not just appalachia"
author: 
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(tidycensus)

# Parameters

# Variable: household income in past 12 months below poverty level;
# white, not hispanic or latino

vars <- 
  c(
    "white_household_below_poverty" = "B17010H_002"
  )

#===============================================================================

# Code
```


Where are the majority of white households living below the poverty line? 

```{r}
all_counties <- 
  get_acs(
    geography = "county",
    variables = vars,
    year = 2018
  ) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  transmute(
    fips = as.integer(GEOID), 
    name = NAME, 
    white_household_below_poverty
  )

all_counties %>%
  arrange(desc(white_household_below_poverty))

all_counties_geo <-
  ussf::boundaries(geography = "county") %>%
  mutate("fips" = as.integer(GEOID)) %>%
  left_join(all_counties, by = "fips") 
  
  
 all_counties_geo %>% 
  filter(white_household_below_poverty < 5000) %>%
  ggplot(aes(fill = white_household_below_poverty, geometry = geometry)) +
  geom_sf(size = 0.01) +
  scale_fill_viridis_c()


```


```{r}
# tried doing dot density plot... wasn't very effective
# dens_whp <-  
#   all_counties_geo %>%
#   mutate(
#     area = (st_area(geometry) %>% 
#       units::set_units(miles^2) %>%
#       as.numeric()),
#     dens = white_household_below_poverty / area
#   )
# 
# 
# dens_whp %>%
#   select(fips, name, dens, white_household_below_poverty, area) %>%
#   arrange(desc(dens))
#   
```



2. When news organizations and political pundits want a picture of white poverty, they should go to Appalachia. 

* Use ACS data to show that there are more poor white people living in suburbs and urban areas than there are in rural areas (or something like that)
* (Cite also: folks like JD Vance and Charles Murray who use words like "Appalachians", "Appalachian whites", and "the white working class".)

NOTE:
* Can do raw number of white people living below poverty line
* Can do population density for certain populations; for instance, number of white people living below the poverty line per square mile -- USSF package gives you back square land miles; did this in the spatial visualization reading; can then take the number and divide by that! (aland variable)

