---
title: "Final report"
author: "Thea Rossman"
date: "`r Sys.Date()`"
output: 
  github_document:
    html: true
---

```{r include = FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidycensus)
library(sf)

  # Parameters
app_counties_race_file <-
  "../data/appalachian_counties_race.csv"
cent_app_counties_race_file <- 
  "../data/central_appalachian_counties_race.csv"
segregation_indices_file <- 
  "../data/appalachian_counties_segregation_indices.csv"
segregation_indices_file_pivoted <- 
  "../data/appalachian_counties_segregation_indices_pivoted.csv"
census_tracts_file <- "../data/appalachian_tracts_race.csv"

  # FIPS codes to filter out when mapping population
alaska_hawaii <- c(2, 15)

  # if any code changes in the future, be sure that this parameter match
  # the data being produced by the scripts
acs5_year <- 2018
  # if any code changes in the future, be sure that this parameter match
  # the data being produced by EDA (tracts measures of segregation)
prop <- 0.05



```

## Introduction

My learning goal with this challenge was to build my comfort working with census data. I used the tidycensus package to analyze estimates from the 5-year American Community Survey (ACS-5) and values from the Decennial Census -- identifying patterns, making comparisons, and identifying potential stories. 

Below, I offer some numbers representing race, segregation, and population demographics in the Appalachian region and the Central Appalachian subregion. 


## Defining Appalachia

I defined Appalachia according to the boundaries set by the Appalachian Regional Commission, available on the ARC website.[^1] I defined Central Appalachia according to boundaries set by the Appalachian Community Fund.[^2]

The code to write the list of counties are in the `scripts` folder, as `write_full_appalachia_counties.R` and `write_central_appalachia_counties.R`. 

I used these FIPS codes to filter data for my analysis. 

```{r}
app_counties_race <- 
  app_counties_race_file %>%
  read_csv(
    col_types = 
      cols(
        county_name = col_character(),
        state_name = col_character(),
        .default = col_integer()
      )
  )

cent_app_counties_race <- 
  cent_app_counties_race_file %>%
  read_csv(
    col_types = 
      cols(
        county_name = col_character(),
        state_name = col_character(),
        .default = col_integer()
      )
  )

cent_app_county_fips <- cent_app_counties_race %>% pull(fips)
app_county_fips <- app_counties_race %>% pull(fips)

```

There are `r nrow(app_counties_race)` counties in this definition of Appalachia, and `r nrow(cent_app_counties_race)` in Central Appalachia. As of `r acs5_year`, the total population of Appalachia is `r app_counties_race %>% pull(total_population) %>% sum() %>% prettyNum(big.mark=",")`, and the population of Central Appalachia is `r cent_app_counties_race %>% pull(total_population) %>% sum() %>% prettyNum(big.mark=",")`.

```{r}
app_counties_race %>%
  select(fips) %>%
  mutate(
    subregion = 
      if_else(fips %in% cent_app_county_fips, 2, 1)
  ) %>%
  full_join(
    ussf::boundaries(geography = "county") %>% 
      transmute(fips = as.integer(GEOID), geometry) %>%
      filter(),
    by = "fips"
  ) %>%
  filter(!(fips %/% 1000 %in% alaska_hawaii)) %>%
  ggplot(aes(fill = subregion, geometry = geometry)) +
  geom_sf(size = 0.1) + 
  geom_sf(
    data = ussf::boundaries(geography = "state") %>%
      transmute(fips = as.integer(GEOID), geometry),
    color = "white", 
    fill = NA, 
    size = 0.2
  ) + 
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(n = 3, name = "Spectral"),
    na.value = "light grey"
  ) +
  theme_void() + 
  guides(fill = "none") + 
  labs(
    title = "Map Showing Appalachia, Central Appalachia",
    caption = "Boundaries from: Appalachian Regional Commission; Appalachian Community Fund"
  )
  
```


## Race, population, and people of color in Appalachia: Demographics

Before digging into contemporary demographics, a historical note is important. The white racial majority that exists in Appalachia today should be understood as the result of centuries-long policies and practices. Colonization, land grants, and the Indian Removal Act decimated Shawnee, Cherokee, Monacan, Seneca, Haudenosaunee, and other societies. Often the first fired and with little union protection, millions of African Americans left Appalachia during migration waves. When the Ku Klux Klan was at its height, many Black families lost property, and terrorism forced others to flee. Many mixed-race people worked to become legally classified as “white” to avoid the full brunt of postbellum Jim Crow.[^3]


```{r}
app_counties_race <- 
  app_counties_race %>%
  mutate(n_nonwhite = total_population - white)

cent_app_counties_race <- 
  cent_app_counties_race %>%
  mutate(n_nonwhite = total_population - white)

n_nonwhite_app <-
  app_counties_race %>%
  pull(n_nonwhite) %>%
  sum()

n_nonwhite_cent_app <- 
  cent_app_counties_race %>%
  pull(n_nonwhite) %>%
  sum()

prop_nonwhite_app <- 
  app_counties_race %>%
  summarize(prop_nonwhite = sum(n_nonwhite) / sum(total_population))

prop_nonwhite_cent_app <- 
  cent_app_counties_race %>%
  summarize(prop_nonwhite = sum(n_nonwhite) / sum(total_population))
  
```

In `r acs5_year`, there were approximately `r n_nonwhite_app %>% prettyNum(big.mark=",")` people of color living in Appalachia, and `r n_nonwhite_cent_app %>% prettyNum(big.mark=",")` people of color living in Central Appalachia. This translates to `r (prop_nonwhite_app*100) %>% round(digits = 1)`% and `r (prop_nonwhite_cent_app*100) %>% round(digits = 1)`% of the regional populations, respectively. 


```{r}
app_counties_race %>%
  pivot_longer(
    cols = c(hispanic_latino, white:multiracial),
    names_to = "race_ethn",
    values_to = "pop"
  ) %>%
  mutate(subregion = "North/South Appalachia") %>%
  bind_rows(
    cent_app_counties_race %>%
      pivot_longer(
        cols = c(hispanic_latino, white:multiracial),
        names_to = "race_ethn",
        values_to = "pop"
      ) %>%
      mutate(subregion = "Central Appalachia")
  ) %>%
  group_by(race_ethn, subregion) %>%
  summarize(
    prop_of_pop = sum(pop) / sum(total_population)
  ) %>%
  filter(race_ethn != "white") %>%
  ggplot(
    aes(
      fct_reorder(race_ethn, -prop_of_pop), 
      prop_of_pop, 
      fill = subregion
    )
  ) + 
  geom_col(position = "dodge") +
  scale_x_discrete(
    labels = function(x) str_replace_all(x, "_", "\n") %>% str_to_title()
  ) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme(
    legend.position = "top", 
    legend.direction = "horizontal"
  ) +
  labs(
    title = "Non-White Populations in Appalachia, Central Appalachia",
    subtitle = "African Americans make up largest non-white group
      Central Appalachia is whiter than region as a whole",
    x = "",
    y = "Percentage of Total \nRegional Population",
    fill = ""
  )
  
```




## Geographic Distribution of African American Population in Appalachia

### By county and subregion

The difference in African American population between Appalachia and Central Appalachia can be explained by the concentration of majority-African American counties in Southern-most Appalachia, particularly in Georgia and Alabama:

```{r}
app_counties_race_geo <- 
  app_counties_race %>%
  left_join(
    (ussf::boundaries("county", projection = "albers") %>% 
       transmute(fips = as.integer(GEOID), geometry)),
    by = "fips"
  )

label_pct <- function(x, pct) {
  if_else(
    x < pct, 
    str_c(as.character(x*100), "%"), 
    str_c(">", as.character(pct*100), "%")
  )
}

state_boundaries <- 
  ussf::boundaries("state", projection = "albers") %>%
  filter(as.integer(GEOID) %in% (app_county_fips %/% 1000))

app_counties_race_geo %>%
  mutate(prop_black = pmin(black / total_population, 0.75)) %>%
  ggplot() +
  geom_sf(
    aes(fill = prop_black, geometry = geometry), 
    size = 0.01
  ) +
  geom_sf(data = state_boundaries, color = "white", fill = NA, size = 0.2) +
  scale_fill_viridis_c(
    breaks = seq(0.05, 0.75, 0.1),
    labels = function(x) label_pct(x, 0.75)
  ) +
  guides(
    fill =
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(size = 12, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  ) +
  labs(
    title = "African American Population in Appalachia",
    subtitle = "Proportion of County Identifying as African American, Non-Hispanic, Alone\n",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
```


Several counties in Alabama and Mississippi are majority-Black: 

```{r}
app_counties_race_b <- 
  app_counties_race %>%
  mutate(prop_black = black / total_population) %>%
  select(county_name, state_name, prop_black, total_population, black) %>%
  arrange(desc(prop_black)) %>%
  filter(prop_black > 0.5)

app_counties_race_b %>%
  knitr::kable()

```

In central Appalachia, there are 3 counties with Black populations over 10%: 

```{r}
cent_app_counties_race_b <- 
  cent_app_counties_race %>%
  mutate(prop_black = black / total_population) %>%
  select(county_name, state_name, prop_black, total_population, black) %>%
  arrange(desc(prop_black)) %>%
  filter(prop_black > 0.1)

cent_app_counties_race_b %>%
  knitr::kable()
```

There are `r nrow(app_counties_race_b %>% filter(prop_black > 0.5))` counties in Appalachia, all in Alabama and Mississippi, with Black-alone populations over 50%, and `r nrow(app_counties_race_b)`, with one in Georgia, with African American populations over 40%. These `r nrow(app_counties_race_b)` counties include a total of `r app_counties_race_b %>% pull(black) %>% sum() %>% prettyNum(big.mark=",")` individuals, or around `r (((app_counties_race_b %>% pull(black) %>% sum()) / (app_counties_race %>% pull(black) %>% sum())) %>% round(digits = 3))*100`% of the African American population in the Appalachian region as a whole. 

When we restrict our attention to Central Appalachia alone, there are only `r nrow(cent_app_counties_race_b)` counties with African-American-alone populations over 10%, with a maximum proportion of `r ((cent_app_counties_race_b %>% pull(prop_black) %>% max()) %>% round(digits = 3))*100`%.

To map the distribution of the Black population of Central Appalachia, we shift our definition to include individuals who identified as African American in combination with some other race(s): 

```{r}
cent_app_counties_race_geo_b <- 
  app_counties_race_geo %>%
  filter(fips %in% cent_app_county_fips) %>%
  left_join(
    get_acs(
      geography = "county",
      year = acs5_year,
      variables = c(black_alone_or_in_comb = "B02009_001")
    ) %>%
      transmute(fips = as.integer(GEOID), name = NAME, variable, estimate) %>%
      pivot_wider(names_from = variable, values_from = estimate),
    by = "fips"
  ) %>%
  mutate(
    prop_black = black_alone_or_in_comb / total_population
  )

cent_app_counties_race_geo_b %>%
  mutate(prop_black = pmin(prop_black, 0.15)) %>%
  ggplot() +
  geom_sf(
    aes(fill = prop_black, geometry = geometry), 
    size = 0.01
  ) +
  geom_sf(
    data = state_boundaries %>% 
      filter(as.integer(GEOID) %in% (cent_app_county_fips %/% 1000)), 
    color = "white", 
    fill = NA, 
    size = 0.2
  ) +
  scale_fill_viridis_c(
    breaks = seq(.01, 0.15, 0.02),
    labels = function(x) label_pct(x, 0.15)
  ) +
  guides(
    fill =
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(size = 12, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  ) +
  labs(
    title = "African American Population in Central Appalachia",
    subtitle = "Proportion of County Identifying as African American, Non-Hispanic
    Alone or in Combination with Other Races\n",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
```

As is clear from the map, there are some counties in Central Appalachia with virtually no Black population: 

```{r}
cent_app_counties_race_geo_b %>%
  mutate(
    name = str_c(county_name, " County, ", state_name)
  ) %>%
  select(name, prop_black, total_population, black_alone_or_in_comb) %>%
  arrange(desc(prop_black)) %>% 
  filter(prop_black < 0.0025) %>%
  arrange(prop_black) %>%
  knitr::kable()
```

Some counties in Central Appalachia with relatively high Black populations are located in metropolitan areas, such as Knox County, Tennessee (Knoxville), Hamilton County, Tennessee (Chattanooga), and Kanawha County, West Virginia (Charleston). However, others are located in smaller counties, such as Henry County/Martinsville City, VA, and Jefferson County, WV. 

```{r}
cent_app_counties_race_geo_b %>%
  mutate(name = str_c(county_name, " County, ", state_name)) %>%
  ggplot(aes(prop_black, total_population)) + 
  geom_point() +
  ggrepel::geom_text_repel(
    aes(label = name),
    data =
      . %>%
      mutate(name = map_chr(name, str_replace, " County", "")) %>%
      filter(prop_black > 0.075),
    size = 3,
    direction = "y",
    hjust = -0.3,
    vjust = 0.2
  ) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Central Appalachian Black Population Spread Across Counties of All Sizes",
    subtitle = "Individuals who self-identified as Black/African American Alone
    or in combination with other races",
    x = "Percent of Residents Self-Identified as Black/African American
    Alone or in Combination with Other Races",
    y = "Total Population of County",
    caption = 
      str_glue(
        "Source: American Community Survey, ", 
        acs5_year
      )
  )
```

The graph above may be somewhat misleading. 

The county in West Virginia with the highest African American population is Gilmer County. Initially, this was surprising to me (I've been to Gilmer County). I then realized that Gilmer County is home to a federal prison, located in Glenville, which, according to a Google search, has an incarcerated population of approximately 1,730 people. 

When I tried retrieving data for the variables on the census that correspond to institutionalized populations, I found that values are suppressed (or there is a bug in the Census API): 

```{r}
gilmer_imprisoned_population <-
  get_acs(
    geography = "county",
    year = acs5_year,
    variables = 
      c(
        institutionalized_pop = "B26101_067",
        institutionalized_pop_black = "B26103B_003",
        institutionalized_pop_multiracial = "B26103F_007"
      ),
    state = "West Virginia",
    county = "Gilmer"
  )

gilmer_imprisoned_population
```

Despite lack of data, my hypothesis from online research is that **Gilmer County is overwhelmingly white, but the imprisoned population living there is majority-Black**.[^4] 

McDowell County, WV and Raligh County, WV are also home to federal prisons, and may show a similar trend. However, **more data would be needed to determine this for certain.** 

In rural areas, including Central Appalachia, the incarceration industry has been sold as a solution to economic decline. More data is needed to determine how this is changing the landscape of Appalachia.[^5]


### By census tract

I also conducted EDA and retreived census data at census tract, block groups, and block geographies. I found that there exist census tracts in Appalachia with extremely high Black populations, which is information that we lose at the county level. 



```{r}
census_tracts_race <- 
  census_tracts_file %>%
  read_csv(
    col_types = 
      cols(
        fips = col_character(), 
        name = col_character(),
        prop_nonwhite = col_double(), 
        .default = col_integer()
      )
  ) %>%
  mutate(prop_black = black / total_population)
```


In the Appalachian region as a whole, there are many census tracts in Alabama and Pennsylvania that are estimated to be nearly 100% Black. 

Note that Allegheny County is in the Pittsburgh area; Jefferson County contains Birmingham; and the University of Alabama is located in Tuscaloosa County. Macon County contains Tuskegee, but is not in a major metropolitan area. 


```{r}
census_tracts_race %>%
  select(name, total_population, black, prop_black) %>%
  filter(prop_black > .95) %>%
  arrange(desc(prop_black)) %>%
  knitr::kable()
```

In central Appalachia, the counties containing Chattanooga (Hamilton) and Knoxville (Knox) contained census tracks over 75% Black: 

```{r}
census_tracts_race %>%
  filter(county_fips %in% cent_app_county_fips) %>%
  select(name, total_population, black, prop_black) %>%
  top_n(10, wt = prop_black) %>%
  arrange(desc(prop_black)) %>%
  knitr::kable()
```

Looking closer at census tracts in Central Appalachia with more than a quarter African American population, we see a relationship between county size and the existence of a census tract with a relatively large Black population: 

```{r}
census_tracts_race %>%
  filter(county_fips %in% cent_app_county_fips) %>%
  drop_na(total_population, prop_black) %>%
  group_by(county_fips) %>%
  mutate(total_pop_county = sum(total_population)) %>%
  ggplot(aes(total_pop_county, prop_nonwhite)) +
  geom_point(alpha = 0.6) +
  ggrepel::geom_text_repel(
    aes(
      label = name %>% 
        str_remove("Census Tract") %>%
        str_remove("County") %>%
        str_remove_all("[:digit:]") %>%
        str_remove_all(",") %>%
        str_remove_all("\\.") %>%
        str_trim() %>%
        str_replace(" ", ",")
    ),
    data = 
      . %>% 
      filter(prop_nonwhite > 0.4) %>%
      distinct(county_fips, .keep_all = TRUE),
    size = 3,
    direction = "both",
    vjust = 0.1,
    hjust = -0.3
  ) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(
    breaks = scales::breaks_width(0.2),
    labels = scales::label_percent()
  ) +
  labs(
    title = "Many, but not all, Majority-Black Census Tracts in Central Appalachia 
    are in Large Counties",
    subtitle = "Labelled counties contain tracts with >40% Black population",
    y = "Prop. self-identified as \nAfrican American, Non-Hispanic Alone",
    x = "Total Population of County",
    caption = 
      str_glue(
        "Source: U.S. Census Bureau, American Community Survey, ",
        acs5_year
      )
  )

```

We also note that one county can show significant range in the African American population of its census tracts.


## Distribution of Hispanic/Latino Population

### By county

Though the counties with the largest Hispanic/Latino proportion, relatively, are clustered in Southern Appalachia, they are clustered in the eastern part of the region, rather than, as with the majority-African American counties, the western part. 

```{r}
app_counties_race_geo %>%
  mutate(
    prop_hisp_latino = pmin(hispanic_latino / total_population, 0.2)
  ) %>%
  ggplot() +
  geom_sf(
    aes(fill = prop_hisp_latino, geometry = geometry), 
    size = 0.01
  ) +
  geom_sf(data = state_boundaries, color = "white", fill = NA, size = 0.2) +
  scale_fill_viridis_c(
    breaks = seq(0, 0.2, 0.05),
    labels = function(x) label_pct(x, pct = 0.2)
  ) +
  guides(
    fill =
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(size = 12, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  ) +
  labs(
    title = "Hispanic/Latino Population in Appalachia",
    subtitle = "Proportion of County Identifying as Hispanic/Latino\n",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
```


```{r}
app_counties_race_geo %>%
  filter(fips %in% cent_app_county_fips) %>%
  mutate(
    prop_hisp_latino = pmin(hispanic_latino / total_population, 0.1)
  ) %>%
  ggplot() +
  geom_sf(
    aes(fill = prop_hisp_latino, geometry = geometry), 
    size = 0.01
  ) +
  geom_sf(
    data = state_boundaries %>% 
      filter(as.integer(GEOID) %in% (cent_app_county_fips %/% 1000)), 
    color = "white", 
    fill = NA, 
    size = 0.2
  ) +
  scale_fill_viridis_c(
    breaks = seq(0, 0.1, 0.02),
    labels = function(x) label_pct(x, 0.1)
  ) +
  guides(
    fill =
      guide_colorbar(
        barheight = 0.5,
        barwidth = 15,
        title = NULL
      )
  ) +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.5, 0.95),
    plot.title = element_text(size = 12, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  ) +
  labs(
    title = "Hispanic/Latino Population in Central Appalachia",
    subtitle = "Proportion of County Identifying as Hispanic and/or Latino\n",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
  
```

```{r}
app_counties_race %>%
  filter(fips %in% cent_app_county_fips) %>%
  mutate(prop_hisp_latino = hispanic_latino / total_population) %>%
  ggplot(aes(total_population, prop_hisp_latino)) +
  geom_hline(yintercept = 0.05, color = "white", size = 2.5) +
  geom_point(alpha = 0.6) +
  ggrepel::geom_text_repel(
    aes(label = name),
    data =
      . %>%
      mutate(name = str_c(county_name, ", ", state_name)) %>%
      filter(prop_hisp_latino > 0.05),
    size = 3,
    direction = "y",
    hjust = -0.3,
    min.segment.length = 0.1
  ) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(
    breaks = seq(0, .12, .02),
    labels = scales::label_percent(accuracy = 1)
  ) +
  labs(
    title = "Most Counties in Central Appalachia 
    with over 5% Hispanic/Latino Population are in Tennessee",
    subtitle = "Counties Vary in Size",
    x = "Total Population of County",
    y = "Percent of County Identified as Hispanic/Latino",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
```

With the exception of Jefferson, WV -- in the eastern panhandle of WV, bordering Washington, DC -- and Henry, Virginia -- in the southernmost part of central/western Virginia, bordering North Carolina -- all counties with more than 5% Hispanic/Latino population are in Tennessee. 

### By census tract

There are a few counties in Georgia and Pennsylvania with census tracts that have a majority Hispanic/Latino population: 

```{r}
census_tracts_race <- 
  census_tracts_race %>%
  mutate(prop_hisp_latino = hispanic_latino / total_population)

census_tracts_race %>%
  select(name, prop_hisp_latino, total_population) %>%
  filter(prop_hisp_latino > 0.5) %>%
  arrange(desc(prop_hisp_latino)) %>%
  knitr::kable()
```

Restricting to Central Appalachia, we see that the census tracts with the ten largest proportion of Hispanic/Latino residents are all in Tennessee, from a total of 6 different counties: 

```{r}
census_tracts_race %>%
  filter(county_fips %in% cent_app_county_fips) %>%
  arrange(desc(prop_hisp_latino)) %>%
  select(name, prop_hisp_latino, total_population) %>%
  top_n(10, prop_hisp_latino) %>%
  knitr::kable()
```


```{r}
app_state_fips <-
  app_counties_race %>%
  transmute(
    state_fips = fips %/% 1000,
    state_name
  ) %>%
  distinct(state_fips, state_name)

census_tracts_race %>%
  mutate(state_fips = county_fips %/% 1000) %>%
  left_join(app_state_fips, by = "state_fips") %>%
  drop_na(state_fips, prop_hisp_latino) %>%
  ggplot(aes(fct_reorder(state_name, prop_hisp_latino), prop_hisp_latino)) +
  geom_hline(yintercept = 0.1, color = "white", size = 2) +
  geom_boxplot() +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    labels = scales::label_percent(),
    limits = c(0, 1.3)
  ) +
  geom_text(
    aes(
      label = name %>% 
        str_remove("Census Tract") %>%
        str_remove("County") %>%
        str_remove_all("[:digit:]") %>%
        str_remove_all(",") %>%
        str_remove_all("\\.") %>%
        str_trim() %>%
        str_replace(" ", ","),
      color = central_app
    ),
    data = 
      . %>% 
      group_by(state_name) %>%
      top_n(1, wt = prop_hisp_latino) %>%
      mutate(central_app = (county_fips %in% cent_app_county_fips)),
    size = 3,
    hjust = -0.1
  ) + 
  coord_flip() +
  scale_color_manual(
    values = c("#8b0000", "#00008b"),
    breaks = c(FALSE, TRUE),
    labels = c("North/South Appalachia", "Central Appalachia")
  ) +
  theme(
    legend.position = "top", 
    legend.direction = "horizontal",
    legend.title = element_blank()
  ) +
  labs(
    title = "Census Tracts by Proportion Hispanic/Latino, by State",
    subtitle = "Showing Census Tracts in Appalachia",
    x = "State",
    y = "Proportion of residents in tract self-identified as Hispanic/Latino",
    caption = 
      str_glue(
        "Source: U.S. Census Bureau, American Community Survey", 
        acs5_year
      )
  )
```


Note that there are counties in every state in Appalachia with at least one census track having over 10% Hispanic/Latino population. 


## Approximate Residential Segregation of African American, Hispanic Populations in Appalachia

The analysis above gives a sense of segregation between counties, as well as some sense that there is significant segregation between census tracts, within a single county. 

The analysis below uses Census tracts to approximate how segregated African American and Hispanic populations are in Appalachia, in the counties within which a significant population of them exists.[^6]

For the purposes of this analysis, "significant" will mean that either the county population is at least `r prop*100`% made up of the group, or contains at least one census tract that is.

(In my EDA, I performed similar analysis using census blocks (decennial census) and block groups. In this report, I am only including segregation metrics calculated via census tracts.[^7]) 

I use two measures to quantify racial segregation[^8]: the isolation index and the dissimilarity index. These are imperfect measurements and should be taken as heuristics only, but they are widely used in sociological literature. 

The isolation index is calculated as: 

\( \frac{
  \sum_{i=1}^n \left( \frac{AA_i}{AA_{total}} \cdot \frac{AA_i}{persions_i} \right) 
  - \frac{AA_{total}}{persons_{total}}}{1 - \frac{AA_{total}}{persons_{total}}} \) 
  
Where \(AA_i\) is the number of African Americans in tract i; \(AA_{total}\) is the number of African Americans in the county as a whole; \(persons_i\) denotes total population in tract i.

The isolation index measures the extent to which African American residents are exposed primarily to one another, calculated through the scaled and normalized proportion of the city area inhabited by the average African American. A higher index (closer to 1) indicates higher isolation. 

And the dissimilarity index is calculated as:

\( \frac{1}{2} \sum_{i=1}^N \mid {\frac{AA_i}{AA_{total}} - \frac{AA_i^c}{AA_{total}^c}} \mid \)

This formula uses the same notation as above; additionally, \(AA_i^c\) is the number of non African Americans in tract i, and \(AA_{total}^c\) is the number of non African Americans in the city.

Dissimilarity measures whether a group is more likely to live in some areas vs. others. If African Americans disproportionately reside in some areas (census tracts) of a county relative to non-African Americans, the dissimilarity index between the two groups is high.

(The indices for the Hispanic/Latino population are calculated equivalently.)

These indices are most useful for comparing across places. 

```{r}
segregation_indices <- 
  segregation_indices_file %>%
  read_csv(
    col_types = 
      cols(
        county_name = col_character(),
        state_name = col_character(),
        aa_dissimilarity_index = col_double(),
        hl_dissimilarity_index = col_double(),
        na_dissimilarity_index = col_double(),
        aa_isolation_index = col_double(),
        hl_isolation_index = col_double(),
        na_isolation_index = col_double(), 
        .default = col_integer()
      )
  )

segregation_indices_pivoted <- 
  segregation_indices_file_pivoted %>%
  read_csv(
    col_types = 
      cols(
        county_name = col_character(),
        state_name = col_character(),
        index_name = col_character(),
        race_ethn = col_character(),
        index_value = col_double(),
        pop_group = col_double(),
        prop_of_population = col_double(),
        .default = col_integer()
      )
  )
```

Below, the indices for each groups are mapped onto Appalachian counties. A few notes from this: 

* First, the counties for which this report does and does not quantify segregation. For each group, these are counties that do not meet the criteria of "significance" defined above. 
  + (In thinking about segregation in the entire Appalachian region, we must also think about segregation between counties, which is somewhat discussed above.) 
* Second, though the comparison between the two indices is not absolute, we can heuristically tell that the dissimilarity indices for both groups are much higher than the isolation indices.
  + This suggests that both Latino and African American Appalachians are much more likely to live in some tracts than others within a county, but are likely to live in tracts with groups of other races. 
* Finally, though there are outliers and some geographic pattern, it is nowhere near as striking as the differences in proportional populations between counties explored above. 


```{r}
counties_segregation_geo <- 
  segregation_indices %>%
  left_join(
    ussf::boundaries(geography = "county") %>%
      transmute(county_fips = as.integer(GEOID), geometry),
    by = "county_fips"
  )


plot_segregation <- function(tibble, group, index) {
  breakwidth <- 
    ((tibble %>% pull(index) %>% max(na.rm = TRUE))/5) %>% round(digits = 2)
  tibble %>%
    ggplot(aes(geometry = geometry, fill = index)) + 
    geom_sf() + 
    geom_sf(data = state_boundaries, color = "white", fill = NA, size = 0.2) +
    scale_fill_viridis_c(
      breaks = scales::breaks_width(breakwidth)
    ) +
    guides(
      fill =
        guide_colorbar(
          barheight = 0.5,
          barwidth = 15,
          title = NULL
        )
    ) +
    theme_void() +
    theme(
      legend.direction = "horizontal",
      legend.position = c(0.5, 0.95),
      plot.title = element_text(size = 12, hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5)
    ) +
    labs(
      title = str_glue(index, " of ", group, " Population"),
      subtitle = 
        str_glue("Counties with at least one census tract having ", 
                 prop*100,
                 "% or more\n",
                 group, 
                 " population, or with overall\n", 
                 group, 
                 " population at least ", 
                 prop*100, 
                 "%."
        )
    )
}

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = aa_dissimilarity_index,
      geometry
    ),
  group = "African American",
  index = "Dissimilarity Index"
)

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = hl_dissimilarity_index,
      geometry
    ),
  group = "Hispanic/Latinx",
  index = "Dissimilarity Index"
)

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = aa_isolation_index,
      geometry
    ),
  group = "African American",
  index = "Isolation Index"
)

plot_segregation(
  counties_segregation_geo %>% 
    select(
      index = hl_isolation_index,
      geometry
    ),
  group = "Hispanic/Latinx",
  index = "Isolation Index"
)

```


```{r}
segregation_indices_pivoted %>%
  filter(race_ethn != "nativeam", prop_of_population > 0.05) %>%
  ggplot(aes(index_value, color = fct_reorder(race_ethn, index_value))) +
  geom_density() +
  facet_grid(cols = vars(index_name)) +
  scale_color_discrete(
    breaks = c("hisplatino", "black"),
    labels = c("Hispanic/\nLatinx", "African American")) + 
  theme(legend.position = "top") +
  labs(
    title = "Measures of Segregation in Appalachia",
    subtitle = "Hispanic/Latino Population Smaller, Less Segregated (by Census Tract)
    Showing Counties Where Group Population Exceeds 5%", 
    x = "Index Value (1 indicates higher segregation)",
    y = "Density (Number of Counties)",
    color = "",
    caption = 
      str_glue(
        "Source: American Community Survey, ",
        acs5_year, 
        "\nSegregation Indices Calculated by Census Tract"
      )
  )
```

It is also worth noting that Hispanic Appalachians are less segregated, by both measures, than African American Appalachians. This may be due to in their relative sizes and/or different, historically rooted experiences of systemic racism in the region.



```{r, fig.asp=1}
segregation_indices_pivoted %>%
  filter(race_ethn == "black") %>%
  ggplot(aes(index_value, color = index_name)) + 
  geom_density() +
  facet_wrap(vars(state_name)) +
  theme(legend.position = "top") +
  labs(
    title = "African American Population: Patterns of Segregation Vary by State",
    subtitle = 
      str_glue(
        "Segregation Indices Calculated For Counties With >",
        as.character(prop*100),
        "% African American/Black, Non-Hispanic, alone population,
        or with at least one census tract having >",
        as.character(prop*100),
        "% African American/Black, Non-Hispanic, alone population."
      ),
    x = "Calculated Index for the County (1 = more segregated)",
    y = "Number of Counties", 
    color = ""
  )

segregation_indices_pivoted %>%
  filter(race_ethn == "hisplatino", state_name != "Maryland") %>%
  ggplot(aes(index_value, color = index_name)) + 
  geom_density() +
  facet_wrap(vars(state_name)) +
  theme(legend.position = "top") +
  labs(
    title = "Hispanic/Latino Population: Patterns of Segregation Vary by State",
    subtitle = 
      str_glue(
        "Segregation Indices Calculated For Counties With >",
        as.character(prop*100),
        "% Hispanic/Latino population,
        or with at least one census tract having >",
        as.character(prop*100),
        "% Hispanic/Latino population."
      ),
    x = "Calculated Index for the County (1 = more segregated)",
    y = "Number of Counties",
    color = "",
    caption = str_glue("Source: American Community Survey, ", acs5_year)
  )
```

Finally, we note that patterns of segregation vary by state (recall that for each state plot, these are only the counties with "significant" populations of Black and/or Hispanic/Latino residents).

For instance -- West Virginia has more highly (>0.5 index) dissimilar counties than Appalachian parts of Georgia, but the two rank similarly in terms of isolation. 


## Conclusion

While many counties and census tracts in Appalachia remain majority-white, there are counties and neighborhoods with significant, and even majority, Black, Hispanic/Latino, Indigenous, and/or Native populations. Unlike at a national scaele, Black people make up the majority of the non-white population in both Appalachia as a whole and Central Appalachia. While some areas in Appalachia are relatively integrated, many remain segregated, both within and between counties and states. Hispanic and Latino Appalachians, due perhaps to their relative size and different experiences of historical racism, tend to be more residentially integrated than Black Appalachians.

The American Community Survey data is abundant and useful, and it offers imperfect but effective estimates for our communities at neighborhood, county, regional, and national scales. There is so much more to do with this dataset!!!


[^1]: A map of the Appalachian region is available on the [ARC website](https://www.arc.gov/images/appregion/AppalachianRegionCountiesMap.pdf). To access FIPS codes, I used supplementary data tables from a study of county economic status, which is available for download [here](https://www.arc.gov/appalachian_region/CountyEconomicStatusandDistressedAreasinAppalachia.asp).

[^2]: Counties are listed and mapped [here](http://www.appalachiancommunityfund.org/where-we-fund/).

[^3]: {A few of many possible sources for this include: *Blacks in Appalachia*, ed. William H. Turner and Edward J. Cabbell (Lexington, KY: University Press of Kentucky, 1985; Anita Puckett, "The Melungeon Identity Movement and the Construction of Appalachian Whiteness," Journal of Linguistic Anthropology 11, no. 1 (June 2001)} 

[^4]: Lynette Monroe describes Gilmer County and its federal prison in a piece for *Blavity*: "Visiting My Boyfriend In A Federal Prison Led Me To Realize Just How Much Black Bodies Are Used As A Solution To Economic Decline," July 19 2018. https://blavity.com/visiting-my-boyfriend-in-a-federal-prison-led-me-to-realize-just-how-much-black-bodies-are-used-as-a-solution-to-economic-decline?category1=community-submitted

[^5]: Links to articles and resources on this: [New Republic](https://newrepublic.com/article/155660/appalachia-coal-mining-mountaintop-removal-prison-fight); [WV Public](https://www.wvpublic.org/post/phantom-promise-how-appalachia-was-sold-prisons-economic-lifeline#stream/0); [Appalachian Prison Book Project](https://appalachianprisonbookproject.org); [Truthout](https://truthout.org/articles/land-of-extraction-how-prison-industry-settled-in-central-appalachia/); [Appalachian Fellows' Blog](http://www.appfellows.org/appfellows-blog/2018/9/6/land-reform-cant-happen-without-the-abolishing-the-prison-industrial-complex-by-lill-prosperino).

[^6]: This [study](https://www.census.gov/prod/2002pubs/censr-3.pdf) from the US Census Bureau uses tracts as the unit of analysis for measuring segregation in metropolitan areas. This [study](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5285374/) used similar measures for rural areas.

[^7]: Hispanic/Latino population numbers are suppressed for blocks and block groups, and census blocks are only available for the decennial census. However, I'm proud of the code I wrote for my EDA to get data for block groups -- map_dfr and functions are so cool!

[^8]: The formulas for these indices come from this [source](https://ignaciomsarmiento.github.io/2017/03/24/Mapping-Segregation-in-Chicago.html), and are duplicated in the sources cited in footnote 8.

